Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-16 scagnostics.Rmd
finished polishing

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of noisy index functions, and @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. On the other hand the global maximum may not be the only projection of interest, and one may want to observe the selected views in the context of the full distribution rather than a static view. These issues are addressed when combining projection pursuit with the grand tour [@CBCH94]. In this case the properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. In addition the tourr package provides the \texttt{search\_geodesic} search function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

\textcolor{red}{\cite{posse95b} discusses the optimization, in particular that for most index functions and optimists results are too local, largely dependent on starting point. Fiedmans suggestion of hybrid stepping search + steepest-ascent still remains fairly local. \cite{posse90} suggested random search instead, which seems to perform better, but did not find pdf of that article, so not so sure what he actually does.}

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers (\cite{diaconis84}, \cite{huber85}, \cite{JS87}, \cite{posse95b}, \cite{hall89}). The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

>>>>

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of A noisy index function. \cite{posse95b} discusses the optimization, in particular that for most index functions and optimal results are too local, largely dependent on starting point. @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. For exploring high-dimensional data, it can be interesting to observe local maxima as well as a global maximum, and thus a hybrid algorithm that still aloows lingering but not being trapped by local maxima is ideal. In addition, being able to visually monitor the optimisation and see the optimal projection in the context of neighbouring projections is useful. This is provided by combining projection pursuit with the grand tour [@CBCH94]. The properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. The tourr package also provides the \texttt{search\_geodesic} function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

<!--
\textcolor{red}{\cite{posse95b} discusses the optimization, in particular that for most index functions and optimists results are too local, largely dependent on starting point. Fiedmans suggestion of hybrid stepping search + steepest-ascent still remains fairly local. \cite{posse90} suggested random search instead, which seems to perform better, but did not find pdf of that article, so not so sure what he actually does.}

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers, e.g. @diaconis84, @huber85, @JS87, @posse95b, @hall89. The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.
